### 2023-01-25 paper.Rmd
included Juli's comments

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother

@@

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity, even though both goals oftentimes coincide. High between-group similarity is for example desirable when designing experimental conditions using different stimulus sets that should be as similar as possible with regard to response-relevant attributes [@lahl2006]. Such stimulus selection tasks are usually conducted on the basis of attribute data and not on pairwise (dis)similarity ratings. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. When attribute values are available, anticlustering approaches that do not require a dissimilarity matrix can be used to partition the data. K-means is probably the best-known clustering objective that can directly be computed on attribute values. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in variance. However, when aiming for overall between-group similarity, neglecting all distribution characteristics other than the mean is misguided; similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, a simulation study shows that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust` [@papenberg2020; @R-anticlust], an extension to the popular statistical programming language R [@R-base].

@@

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all observations belonging the $k$'th cluster $C_k$ ($k = 1, \ldots, K$):

@@

Note there is a direct connection between the $\mathit{SSE}$ and the location of the cluster centroids [@spath1986]. If the $\mathit{SSE}$ is minimal---which is the goal of k-means *clustering*---, the centroid values $\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as possible from the overall data centroid $\overline{\mathbf{x}} = (\overline{\mathbf{x}}_1, \overline{\mathbf{x}}_2, \ldots, \overline{\mathbf{x}}_P$), where $\overline{\mathbf{x}}_p$ is the overall mean value on the $p$th attribute:

@@

To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable; the new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we can define the criterion $\mathit{SSE_{Var}}$ reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. This simplification leads to the *unweighted sum approach* for k-plus anticlustering: When defining $z_{i.} = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, y_{iP})$, the bicriterion $\mathit{SSE_{kplus}}$ can be computed as

$$
\mathit{SSE_{kplus}} =
  \sum\limits_{j=1}^{2 P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2
$$.

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending them to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Hence, using the unweighted sum approach, k-plus anticlustering leaves the original k-means criterion unchanged. Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardized before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

The three k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means; (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups; (iii) a simple random allocation of elements to groups. All anticlustering methods used a $z$-standardization of the data sets as a preprocessing step.

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations had to be averaged across pairs of features instead of single features.

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus terms in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering was capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria aggravated fulfilling each single criterion: k-plus skew/kurtosis and k-plus correlation performed worse at minimizing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$ than standard k-plus. However, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, k-plus skew/kurtosis was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

@@

In the first example, I use norming data for the OASIS image data set that is freely available online [@kurdi2017introducing]. Kurdi et al. assembled 900 open-access color images and collected ratings on two affective dimensions: arousal and valence. @brielmann2019intense collected an additional rating dimension by measuring how the same 900 images were rated with regard to their beauty. In my application, I used all three features; the corresponding data is available from Github.[^githubbeauty]

@@

The function call takes about 10 seconds on a contemporary personal computer; the run time may vary depending on how many iterations are needed to find the local maximum. The function `kplus_anticlustering()` has four boolean arguments to specify whether variance, skewness, kurtosis and covariances should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness, kurtosis and convariance have to be spefically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features are standardized before the optimization starts; in particular, this option enforces that all k-plus criteria receive the same weight during the optimization process. This behaviour can be adjusted using the boolean argument `standardize`. The documentation of the `anticlust` package also provides detailed information on how to use the function `kplus_anticlustering()`.

Table 2 shows the results of the anticlustering application by listing the descriptive statistics (means, standard deviations and skewness) for each of the 9 groups and for each of the 3 features. Table 2 also contains the results after applying k-means anticlustering and diversity anticlustering. It is shown that---up to two decimals---k-plus anticlustering perfectly matched all groups with regard to the features' means and standard deviations, and the skewness was also very evenly matched. K-means anticlustering also perfectly matched the mean values, but showed decreased performance with regard to similarity in standard deviations skewness in particular. Diversity anticlustering was well-suited to match means, standard deviations and skew, but was slightly outperformed by k-plus. The latter result is not surprising because---unlike k-plus anticlustering---diversity anticlustering does not directly maximize similarity with regard to these criteria, but instead maximizes within-group heterogeneity.

@@

A simulation study and examples on real norming data showed that the k-plus criterion is well suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be aggravated in small data sets. In practice, users can try out different objectives and see which specification fulfils their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are subpar, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

@@

The present paper focused on the new k-plus criterion for anticlustering, but only superficially considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications showed that the local maximum search, maximizing the unweighted k-plus sum, yielded satisfying results that probably suffice in most applications. It is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (see Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives, i.e., the classical k-means criterion and the diversity. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.

[^clusterediting]: In their study, they used the term anticluster editing to refer to the maximization of the diversity criterion. This is because the diversity criterion is minimized as part of the cluster editing method [@zahn1964; @shamir2004vf].

>>>>

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
  - |
    \makeatletter
    \renewcommand*\env@matrix[1][*\c@MaxMatrixCols c]{%
      \hskip -\arraycolsep
      \let\@ifnextchar\new@ifnextchar
      \array{#1}}
    \makeatother

@@

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity, even though both goals oftentimes coincide. High between-group similarity is for example desirable when designing experimental conditions using different stimulus sets that should be as similar as possible with regard to response-relevant attributes [@lahl2006]. Such stimulus selection tasks are usually conducted on the basis of attribute data and not on pairwise (dis)similarity ratings. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. When attribute values are available, anticlustering approaches that do not require a dissimilarity matrix[^dissimilarityfromfeatures] can be used to partition the data. K-means is probably the best-known clustering objective that can directly be computed on attribute values. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in variance. However, when aiming for overall between-group similarity, neglecting all distribution characteristics other than the mean is misguided; similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion such as the diversity to obtain a partitioning.

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, a simulation study shows that k-plus anticlustering is well-suited to create groups having minimal differences according to multiple distribution characteristics. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust` [@papenberg2020; @R-anticlust], an extension to the popular statistical programming language R [@R-base].

@@

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, where each entry is the mean value of one of the $P$ attributes, computed across all observations belonging to the $k$'th cluster $C_k$ ($k = 1, \ldots, K$):

@@

Note that there is a direct connection between the $\mathit{SSE}$ and the location of the cluster centroids [@spath1986]. If the $\mathit{SSE}$ is minimal---which is the goal of k-means *clustering*---, the centroid values $\overline{\mathbf{x}}^{(k)} (k = 1, \ldots, K)$ are as far away as possible from the overall data centroid $\overline{\mathbf{x}} = (\overline{\mathbf{x}}_1, \overline{\mathbf{x}}_2, \ldots, \overline{\mathbf{x}}_P$), where $\overline{\mathbf{x}}_p$ is the overall mean value on the $p$th attribute:

@@

To obtain an adaptation of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable; the new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we can define the criterion $\mathit{SSE_{Var}}$ reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. This simplification leads to the *unweighted sum approach* for k-plus anticlustering: When defining a new data matrix $\mathbf{Z}_{N \times 2P} = \{z_{ij}\}_{N \times 2P}$ as the augmentation of $\mathbf{X}_{N \times P}$ with $\mathbf{Y}_{N \times P}$, i.e.

$$
  \mathbf{Z}_{N \times 2P} =
  \left[ {\begin{array}{cccc|cccc}
    x_{11} & x_{12} & \cdots & x_{1P} & y_{11} & y_{12} & \cdots & y_{1P}\\
    x_{21} & x_{22} & \cdots & x_{2P} & y_{21} & y_{22} & \cdots & y_{2P}\\
    \vdots & \vdots & \ddots & \vdots & \vdots & \vdots & \ddots & \vdots\\
    x_{N1} & x_{N2} & \cdots & x_{NP} & y_{N1} & y_{N2} & \cdots & y_{NP}\\
  \end{array} } \right]
$$

and $z_{i.} = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, y_{iP})$, the bicriterion $\mathit{SSE_{kplus}}$ can be computed as

$$
\mathit{SSE_{kplus}} =
  \sum\limits_{j=1}^{2 P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2.
$$

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original values $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Hence, using the unweighted sum approach, k-plus anticlustering leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardized before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

The three k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means; (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups; (iii) a simple random allocation of elements to groups.

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I analyzed the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations had to be averaged across pairs of features instead of single features.

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that k-plus anticlustering is capable of addressing multiple objectives at the same time---especially when $N$ increases. However, while k-plus anticlustering was capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria made it more difficult to fulfill each single criterion: k-plus skew/kurtosis and k-plus correlation performed worse at minimizing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$ than standard k-plus. However, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, k-plus skew/kurtosis was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

@@

In the first example, I use norming data for the OASIS image data set that is freely available online [@kurdi2017introducing]. Kurdi et al. assembled 900 open-access color images and collected ratings on two affective dimensions: arousal and valence. @brielmann2019intense collected additional norming data by measuring how the same 900 images were rated with regard to their beauty. In my application, I used all three features; the corresponding data is available from Github.[^githubbeauty]

@@

The function call takes about 10 seconds on a contemporary personal computer. The function `kplus_anticlustering()` has four boolean arguments to specify whether variance, skewness, kurtosis and covariances should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness, kurtosis and convariance have to be spefically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features are standardized before the optimization starts; in particular, this option enforces that all k-plus criteria receive the same weight during the optimization process. This behaviour can be adjusted using the boolean argument `standardize`. The documentation of the `anticlust` package also provides detailed information on how to use the function `kplus_anticlustering()`.

Table 2 shows the results of the anticlustering application by listing the descriptive statistics (means, standard deviations and skewness) for each of the 9 groups and for each of the 3 features. Table 2 also contains the results after applying k-means anticlustering and diversity anticlustering. It is shown that---up to two decimals---k-plus anticlustering perfectly matched all groups with regard to the features' means and standard deviations, and the skewness was also very evenly matched. K-means anticlustering also perfectly matched the mean values, but showed decreased performance with regard to similarity in standard deviations, and skewness in particular. Diversity anticlustering was well-suited to match means, standard deviations and skew, but was slightly outperformed by k-plus. The latter result is not surprising because---unlike k-plus anticlustering---diversity anticlustering does not directly maximize similarity with regard to these criteria, but instead maximizes within-group heterogeneity.

@@

A simulation study and examples on real norming data showed that the k-plus criterion is well-suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be challenging in small data sets. In practice, users can try out different objectives and see which specification fulfills their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are subpar, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

@@

The present paper focused on the new k-plus criterion for anticlustering, but only superficially considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications showed that the local maximum search, maximizing the unweighted k-plus sum, yielded satisfying results that probably suffice in most applications. It is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives, i.e., the classical k-means criterion and the diversity. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.

[^clusterediting]: In their study, they used the term anticluster editing to refer to the maximization of the diversity criterion. This is because the diversity criterion is minimized as part of the cluster editing method [@shamir2004vf].
